---
title: Vitamin D - COVID-19 MR (redo)
output: pdf_document
params:
  outcome_gwas: COVID-A2
  exposure_name: Vitamin D
  prefix: VD_A2
  pop: !r c("CEU", "TSI", "GBR", "IBS")
  LDLink_token: !r tryCatch(glue::trim(readChar("../LD_LINK_TOKEN", 20)), warning=function(cond) { message("The file LD_LINK_TOKEN should contain an access token to the NCI API. if you don't have a token go here:\n https://ldlink.nci.nih.gov/?tab=apiaccess"); return(""); })
  near_half_threshold: 0.08
  exposure_gwas_file: ../raw_data/1-s2.0-S0002929720300173-mmc2.xlsx
  exposure_pvalue_threshold: !r as.numeric("5e-5") 
  MAF_rare_threshold: 0.01
  prune_r2_threshold: 0.05
  skip_LD_api: !r TRUE
  outcome_file: ../raw_data/COVID19_HGI_A2_ALL_leave_23andme_20201020.b37.txt.gz
  results_dir: ../derived_data
  use_cache: TRUE
---

```{r setup}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(cache = params$use_cache)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(comment = "")

```

```{r libraries and setup}
library(tidyr)
library(dplyr)
library(plyr)
library(readxl)
library(glue)
library(magrittr)
library(ggplot2)
library(validate)

source("utils.R")

for (i in names(params)) { 
  if (i != "LDLink_token") 
    print(glue("{i} = {params[i]}")) 
  else 
    print(glue("{i} = [redacted]"))
}

```


## Extract data from VitaminD GWAS
Since the input data is given to us as an excel spreadsheet, we have to extract the information from it
"manually". In order to avoid copy-paste errors, we extract the information using the readxl library. 

```{r extract excelsheet data, verbose=TRUE}
data = read_excel(params$exposure_gwas_file, 
                  sheet = "Table S2", 
                  skip = 4, 
                  .name_repair = "minimal") %>% 
  subset(select = c(1:5, 13:17))
```

```{r debug1,verbose=TRUE}
first_blank = min(which(is.na(data$RSID)))
# use the first blank line as an indication of the end of the data, since the dictionary comes after
if (length(first_blank) == 1) {
  data = data[1:min(first_blank - 1, nrow(data)),]
}
```

```{r debug2,verbose=TRUE}
names(data)[8] = "beta"
names(data)[1] = "rsid"
data %<>% 
  mutate_at(.vars = c("EAF", "beta", "SE", "P"), as.numeric) %>% 
  mutate(POS = BP, CHR = gsub("^chr", "", CHR), BP = NULL)
```


Make sure that the exposure data validates as a gwas
```{r Validate exposure data}
summary(confront(data, gwas_rules))
```

```{r debug4,verbose=TRUE}
significant_snps <- filter_and_write_exposure_data(data, 
                                                   location_prefix = glue("{params$results_dir}/{params$prefix}_"), 
                                                   params$exposure_pvalue_threshold, 
                                                   params$MAF_rare_threshold)
```
Read `r nrow(data)` entries from `r params$exposure_gwas_file`

### Outcome  

## extract relevant SNPs
```{r extract Snps}

map_outcome_data <- function(x) {
  
  mutate(x, 
         beta = all_inv_var_meta_beta, 
         P = all_inv_var_meta_p, 
         EA = ALT, 
         NEA = REF, 
         EAF = all_meta_AF, 
         SE = all_inv_var_meta_sebeta) %>% 
      subset(select = required_headers)
}
 
  outcome_data <- extract_snps_from_bgzip(outcome = params$outcome_file, snps = significant_snps) %>% map_outcome_data()
  
  print(glue("There are {nrow(outcome_data)} SNPs in the outcome file {params$outcome_file}"))
```


Make sure that the outcome data validates as a gwas
```{r Validate outcome data}
summary(confront(outcome_data, gwas_rules))
```
```{r put output file name into variable}
rsIdsFromDBSNP = glue("{params$results_dir}/{params$prefix}_rsIdsFromdbSnp.txt")

```
## find names for un-named snps
```{r find_unnamed_snps, cache=TRUE, extra.cache=tools::md5sum(dir("../derived_data/","rs[0-9]*.txt", full.names = TRUE))}
## this can take time and hits the API multiple times....
print(glue("There are {nrow(outcome_data %>% subset(is.na(rsid)))} SNPs in the outcome data that are missing their rsid: "))
print(outcome_data %>% subset(is.na(rsid), select = c(CHR, POS, EA, NEA)))

withIds <- get_unknown_rsids_from_locus(outcome_data)
print(glue("{nrow(withIds)} SNPs rsids were found:"))
print(glue("{withIds$rsid}"))

write.table(withIds, rsIdsFromDBSNP, quote = F, sep = '\t', row.names = F)
```

```{r post-cache read variable}
withIds <- read.table(rsIdsFromDBSNP, sep = '\t', header = T)

```

## merge into other data
```{r merge found snps into data}
outcome_with_rsids <- merge_rsids_into_gwas(outcome_data, withIds)
  
print(glue("After merging the results, there are {nrow(outcome_with_rsids%>%subset(is.na(rsid)))} variants lacking rsid, and {nrow(outcome_with_rsids)} variants overall."))
```

```{r merge exposure and outcome}
merged_exp_and_out <- merge(significant_snps, suffixes = c(".exp", ".out"),
                          outcome_with_rsids, 
                          by = c("rsid", "CHR", "POS"), 
                          all.x = T,
                          )

print(glue("After merging Exposure and Outcome, there are {nrow(merged_exp_and_out%>%subset(is.na(beta.out)))} variants lacking 'beta.out', and {nrow(merged_exp_and_out)} variants overall."))
```

```{r get proxies, cache=TRUE}

##takes time (hits the LDLink API)
proxies_raw <- get_proxies(subset(merged_exp_and_out, is.na(beta.out))$rsid, 
                           params$LDLink_token, 
                           population = params$pop, 
                           results_dir = params$results_dir, 
                           skip_api = params$skip_LD_api)

print(glue("Found {nrow(proxies_raw)} proxies for {length( unique(proxies_raw$query_rsid))} variants, out of {length(which(is.na(merged_exp_and_out$EAF.out)))} requested"))
```

Subset to proxies that are in the outcome first, since we need to find out which proxies are present in the outcome.
```{r use proxies}
proxies_in_outcome <- extract_snps_from_bgzip(params$outcome_file, snps = proxies_raw) %>% map_outcome_data()

print(glue("need to look for rsid for {nrow(subset(proxies_in_outcome, is.na(rsid)))} "))

rsIdsFromDBSNP_proxies=glue("{params$results_dir}/{params$prefix}_proxy_rsIdsFromdbSnp.txt")
```


```{r find proxies_missing_rsids, cache=TRUE}
proxies_in_outcome_with_missing_ids <- get_unknown_rsids_from_locus(proxies_in_outcome)

write.table(x = proxies_in_outcome_with_missing_ids, file = rsIdsFromDBSNP_proxies, quote = F, sep = '\t', row.names = F)
  
```

```{r merge with rest of outcome}
proxies_in_outcome_with_missing_ids <- read.table(file = rsIdsFromDBSNP_proxies,  sep = '\t')

proxies_in_outcome_with_rsids <- merge_rsids_into_gwas(proxies_in_outcome, proxies_in_outcome_with_missing_ids)

proxies <- proxies_raw %>% subset(rsid %in% proxies_in_outcome_with_rsids$rsid)

```
found rsids for `r #nrow(proxies_in_outcome_with_missing_ids)` SNPs


There are `r #nrow(proxies)` proxies 
(of `r #nrow(proxies_raw)`) in the outcome.

Merge proxies with exposure...
```{r merge proxies and exposure}
exposure_and_proxies <- merge(proxies, significant_snps, by.x = c("query_rsid","CHR"), by.y = c("rsid","CHR"), suffixes = c(".proxy",".exp"))
```
... `r nrow(exposure_and_proxies)` proxies were successfully merged with the exposure.

Remove chimeric snps and then the smallest P
```{r, remove chimeric SNPs}
proxies_with_exposure <- exposure_and_proxies %>% 
  mutate(maf_near_half = abs(EAF.exp - 0.5) <= params$near_half_threshold, 
         distance_to_query = abs(POS.exp - POS.proxy)) %>%
  subset( !(Alleles %in% chimeric & maf_near_half)) %>%
  group_by(query_rsid) %>%
  slice_max(order_by = R2, n = 1, with_ties = TRUE) %>%
  slice_min(order_by = distance_to_query, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(maf_near_half = NULL, 
         distance_to_query = NULL) 
```
After removing ambiguous snps and choosing the most linked SNP for every query, `r nrow(proxies_with_exposure)` proxies remain.

```{r Check alleles, eval=FALSE}
with(proxies_with_exposure, Alleles == glue("({EA}/{NEA})") | Alleles == glue("({NEA}/{EA})"))
subset(proxies_with_exposure, select = c(Alleles, Correlated_Alleles, rsid, EA, NEA))
```


```{r fix-proxy-alleles}

fixed_proxies_with_exposure <- adply(proxies_with_exposure, 1, function(x) {
  replace_alleles(list(x$EA, x$NEA), x$Correlated_Alleles) %>%
    {
  data.frame(
    EA = .[1],
    NEA = .[2]
    )}
  })

if (!all(with(fixed_proxies_with_exposure, Alleles == glue("({EA}/{NEA})") | Alleles == glue("({NEA}/{EA})")))) {
  warn("there was a problem fixing the alleles")
}

```

```{r merge proxy rows with outcome and combine with rest}

merged_with_proxies <- merge(proxies_in_outcome_with_rsids,
                             fixed_proxies_with_exposure,
                             suffixes = c(".out",".exp"),
                             by.x = c("rsid", "CHR"),
                             by.y = c("rsid", "CHR")) %>%
  select(c(-POS.proxy, -POS.exp, -R2, -Correlated_Alleles, -Alleles, -Locus,))

merged_and_combined <- rbind(merged_with_proxies, mutate(merged_exp_and_out, query_rsid = NA), stringsAsFactors = FALSE) %>% subset(!is.na(EAF.out))
```

Total number of SNPs for use in 2-Sample MR (prior to pruning): `r nrow(merged_and_combined)`



Now we need to make sure that we do not have SNPs in high LD next to each other.
Pruning using populations: [`r glue_collapse(glue("{params$pop}"),sep=", ")`]
```{r prune_snps, cache = TRUE}

ld_pairs_raw <- get_LD_pairs(merged_and_combined, population = params$pop, token = params$LDLink_token)

unique_rsids <- with(ld_pairs_raw, unique(c(as.character(RS_number), as.character(variable))))
ld_pairs <- ld_pairs_raw %>% mutate(
  variable = factor(variable, levels = unique_rsids),
  RS_number = factor(RS_number, levels = unique_rsids))

p = ggplot(ld_pairs) +  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + geom_bin2d(stat = 'identity', aes(x = RS_number, y = variable))
    
show(p + aes(fill = value) + labs(fill = expression(R^2), parse = TRUE))
show(p + aes(fill = value > params$prune_r2_threshold) + labs(fill = expression(R^2 > threshold), parse = TRUE))


```

```{r }
list_result <- prune_snps(merged_and_combined, ld_pairs, params$prune_r2_threshold)


pruned_combined_snps <- list_result$rsids
removed_rsids <- list_result$removed_rsid

ggplot(ld_pairs) +  
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  geom_bin2d(stat = 'identity', aes(x = RS_number, y = variable)) +
  aes(fill = value > params$prune_r2_threshold, alpha = RS_number %in% removed_rsids | variable %in% removed_rsids) + 
  labs(fill = expression(R^2>threshold), alpha = "variant removed", parse = TRUE) + 
  scale_alpha_manual(values = c(0.3, 1), breaks = c(TRUE, FALSE))
 

write.table(pruned_combined_snps, file = glue("{params$results_dir}/{params$prefix}_SNPsFor2SMR.txt"), sep = '\t', quote = F, row.names = F)
```
Removed `r length(removed_rsids)` snps due to high LD: `r glue_collapse(glue("{removed_rsids} "))`
Total number of SNPs for use in 2-Sample MR (pruned): `r nrow(pruned_combined_snps)`

```{r get MR results}
results <- get_2smr_results(glue("{params$results_dir}/{params$prefix}_SNPsFor2SMR.txt"))
```

```{r show results}

for (p in results) {
  show(p)
}
save(results, file = glue("{params$results_dir}/{params$prefix}_2SMR_results.RData"))

```

```{r compare with Guillaumes results, eval=FALSE} 
rsids=c("rs6698680", "rs3750296", "rs7519574", "rs56044892", "rs2934744", "rs7528419", "rs3768013", "rs115045402", "rs144613541", "rs61816761", "rs11264360", "rs867772", "rs12997242", "rs11127048", "rs6724965", "rs7569755", "rs1047891", "rs2011425", "rs1972994", "rs6438900", "rs6773343", "rs78649910", "rs7699711", "rs11723621", "rs58073039", "rs28364331", "rs1229984", "rs7718395", "rs3822868", "rs111529171", "rs1011468", "rs1858889", "rs804280", "rs34726834", "rs7828742", "rs10818769", "rs532436 (proxy: rs635634)", "rs10887718", "rs10832218", "rs577185477", "rs10832289", "rs188480917 (proxy: rs576128895) ", "rs117576073", "rs523583", "rs12803256", "rs200454003 (proxy: rs4245442)", "rs10793129", "rs1149605", "rs964184", "rs2847500", "rs12317268", "rs9668081", "rs10859995", "rs8018720", "rs261291", "rs1800588", "rs17765311", "rs62007299", "rs8063706", "rs77924615", "rs1800775", "rs2909218", "rs8091117", "rs2037511", "rs57631352", "rs73015021", "rs10500209", "rs58542926", "rs3814995", "rs1065853", "rs157595", "rs112285002 (proxy: rs296386)", "rs10426", "rs8103262", "rs6123359", "rs6127099", "rs2585442", "rs2762942", "rs2229742", "rs2074735", "rs960596")%>%gsub(pattern = " .*",replacement = "")

###investigate differences:

setdiff(pruned_combined_snps$rsid,rsids)
setdiff(rsids, pruned_combined_snps$rsid)

rsids_differences=rbind(data.frame(rsid=setdiff(pruned_combined_snps$rsid,rsids), set="Y"),
                        data.frame(rsid=setdiff(rsids,pruned_combined_snps$rsid), set="G"))

subset(proxies, rsid %in% rsids_differences$rsid)


ld_pairs_raw <- get_LD_pairs(rsids_differences$rsid, population = params$pop, token = params$LDLink_token)


subset(pairs, RS_number %in% setdiff(pruned_combined_snps$rsid,rsids) || RS_number %in% setdiff(rsids,pruned_combined_snps$rsid))

```



